Automatic Identification and Annotation of Parallel Tasks in Structured Programs

### 2017-12-06 docs/ACM/paper.tex
Adding Section 2.1 to the paper.

This paper describes \Taskminer, a compiler that mines task parallelism
from programs.
In order to fulfil this goal, \Taskminer{} solves many challenges.
First, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:identification}).
Second, it determines symbolic bounds to all the memory blocks accessed within
those regions (Section~\ref{sub:symb}).
Third, it extracts parameters from the code to estimate the profitability of
tasks, given the cost to create them.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}) and limit their recursion
depth (Section~\ref{sub:rec}).
Additionally, \Taskminer{} determines which program variables need to be replicated at
each new task, or shared among them (Section~\ref{sub:variance}).
Finally, it maps all this information back into source code, producing annotations
that programmers can read (Section~\ref{sub:ir}).
This work demonstrates the design and the implementation of these technologies, 
as well as their benefit.

@@

\section{Overview}
\label{sec:ovf}

The automatic insertion of effective OpenMP annotations into C/C++ programs
required us to deal with a number of challenges.
If left unsolved, these challenges would restrict our interventions in programs
to trivial annotations, which would hardly be of any use.
In the rest of this section, we discuss the most important challenges, using
this discussion as a motivation to the developments that follow in
Section~\ref{sec:sol}.

@@

We introduce Challenge~\ref{ch:cutoff} by quoting Duran {\em et al.}:
``{\em In task parallel languages, an important factor for achieving a good
performance is the use of a cut-off technique to reduce the number of tasks
created}"~\cite{Duran08}.
This observation is particularly true in the context of recursive, fine-grain,
tasks, as we analyze in Section~\ref{sub:rec}.
Figure~\ref{fig:ex_cutoff} provides an example.
To place a limit on the number of tasks simultaneously in flight, we associate
the invocation of recursive functions annotated with task pragmas with a
counter -- \textsf{taskminer\_depth\_cutoff} in Figure~\ref{fig:ex_cutoff}.
The guard in line 7 ensures that we never exceed \textsf{DEPTH\_CUTOFF}, a
predetermined threshold.
This example, together with Figure~\ref{fig:ex_Regions}, lets us emphasize that
the code generation algorithms presented in this paper are
parameterized by constants such as \textsf{DEPTH\_CUTOFF}, or
\textsf{WORK\_CUTOFF} in Figure~\ref{fig:ex_Regions}. Although these
have default estimates, we provide them as parameters to be set at will.

>>>>

This paper describes \Taskminer, a source-to-source compiler that exposes task
parallelism in C/C++ programs.
To fulfil this goal, \Taskminer{} solves different challenges.
First, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:identification}).
Second, it determines symbolic bounds to the memory blocks accessed within
those regions (Section~\ref{sub:symb}).
Third, it extracts parameters from the code to estimate when it is profitable to
create tasks.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}) and limit their recursion
depth (Section~\ref{sub:rec}).
Fourth, \Taskminer{} determines which program variables need to be 
replicated in new tasks, or shared among them (Section~\ref{sub:variance}).
Finally, it maps all this information back into source code, producing
readable annotations (Section~\ref{sub:ir}).

@@

\section{Advantages and Difficulties}
\label{sec:ovf}

The use of OpenMP pragmas as the concrete means to explore, without human
intervention, task parallelism in programs presents difficulties and
advantages.
Such challenges and benefits have guided the design and implementation of the
techniques that we discuss in Section~\ref{sec:sol}.
In the rest of this section, we discuss them.

\subsection{The Advantages}
\label{sub:adv}

The main advantage of using OpenMP to split a program into parallel tasks is
its runtime.
This runtime lets us circumvent the shortcomings that pointer aliasing
impose on the automatic parallelization of code.
Aliasing -- the possibility of two pointers dereferencing the same
memory region -- either prevents parallelism altogether, or forces compilers to
resort to complex runtime checks to identify it
dynamically~\cite{Alves15,Rus02}.
The OpenMP runtime shields us from having to implement such checks, because
it already checks for dependencies among tasks, and dispatches them in a
correct order~\cite{LaGrone11}.
An OpenMP task descriptor contains the task's status (running, waiting) and a
list of memory addresses the task registered as input arguments.

The process of checking for dependencies can be as simple as checking if an
incoming task in the system accesses a memory address already in use by a
previous task in flight.
It can also be complex, involving checks over long ranges of addresses
(in case the runtime system supports dependence checking for memory ranges).
This process can also include memory labelling and renaming (in case the
runtime system is able to remove false dependencies).
Regardless of its capacity -- which is not the same across every OpenMP
implementation -- the runtime system will represent dependencies using a task
dependence graph (TDG)~\cite{Duran08}.
In this directed acyclic graph, nodes denote tasks and edges represent
dependences between them.
Tasks are dispatched for execution according to a topological ordering of this
graph~\cite{Planas15}.

This runtime support allows the parallelization of irregular applications,
such as programs that traverse data-structures formed by a mesh of pointers.
As an example, Figure~\ref{fig:book_filter} shows an application that
finds patterns in lines of a book.
The book is given as an array of pointers; each pointer leads to a
string representing a --potentially-- different line.
The natural parallelization of this program consists in firing off a task
to process each line.
Figure~\ref{fig:book_filter} has been annotated automatically by the \Taskminer.
We emphasize that neither the runtime checks of Alves
{\em et al.}~\cite{Alves15} or Rus {\em et al.}~\cite{Rus02}, nor a
context and flow sensitive alias analysis, \`{a} la Whaley and
Lam~\cite{Whaley04} would be able to ensure the automatic parallelization of
this program.

\subsection{The Difficulties}
\label{sub:dif}

The automatic insertion of effective OpenMP annotations into C/C++ programs
required us to deal with a number of challenges.
If left unsolved, these challenges would restrict our interventions in programs
to trivial annotations, which would hardly be of any use.
Our first challenge is inherent to any automatic parallelization system.

@@

We introduce Challenge~\ref{ch:cutoff} by quoting Duran {\em et al.}:
``{\em In task parallel languages, an important factor for achieving a good
performance is the use of a cut-off technique to reduce the number of tasks
created}"~\cite{Duran08b}.
This observation is particularly true in the context of recursive, fine-grain,
tasks, as we analyze in Section~\ref{sub:rec}.
Figure~\ref{fig:ex_cutoff} provides an example.
To place a limit on the number of tasks simultaneously in flight, we associate
the invocation of recursive functions annotated with task pragmas with a
counter -- \textsf{taskminer\_depth\_cutoff} in Figure~\ref{fig:ex_cutoff}.
The guard in line 7 ensures that we never exceed \textsf{DEPTH\_CUTOFF}, a
predetermined threshold.
This example, together with Figure~\ref{fig:ex_Regions}, lets us emphasize that
the code generation algorithms presented in this paper are
parameterized by constants such as \textsf{DEPTH\_CUTOFF}, or
\textsf{WORK\_CUTOFF} in Figure~\ref{fig:ex_Regions}. Although these
have default estimates, we provide them as parameters to be set at will.
